Narrative

### 2023-05-21 manuscript/methods.qmd
method: improved fisher z discussion

We use correlations as our effect size type in these simulations because correcting them for unreliability is straightforward and because they are the effect size type most commonly corrected in practice. {see also argument in Field that all effect size types are pretty similar} However, unreliability attenuates all types of effect sizes and all can be corrected, either directly (standardized mean differences; Wiernick & Dahlke, 2020) or by first transforming them into correlations (with associated assumptions).

Perhaps the most common way to meta-analyze correlations is to first transform them into Fisher's z correlations (Borenstein et al., 2009, p. 41 - 43). The Fisher z transformation is the inverse hyperbolic tangent of the product moment correlation and can be computed as $equation$ and its variance approximated as $equation2$. There is some controversy in the use of the Fisher z transformation for meta-analysis (for a good summary see Brannick et al., 2019, p. 492), in particular among proponents of Hunter and Schmidt's (2015) meta-anlytic models.

We perform simulations both on product-moment correlations and Fisher z transformed correlations. For our purpose of estimating heterogeneity, Fisher z correlations have two advantages: 1) their sampling variance is not dependent on the size of the correlation, 2) they are not bounded to {-1, 1}. However, they also have two disadvantages: 1) that the heterogeneity estimate is measured on the Fisher z scale, which makes it hard to interpret. A workaround is to compute a prediction interval and convert this interval onto the product-moment scale (REF) for an indirect estimate of the heterogeneity. Second, adding measurement error to the Fisher z correlations requires first transforming them into product-moment correlations, adding measurement error, and then backtransforming to Fisher z for meta-analysis.

The primary advantage to using product-moment correlations in the meta-analysis when interested in heterogeneity is that values on this scale are more interpretable that on the Fisher z scale. The disadvantage is that the sampling variance is dependent on the effect size since it is computed as $equation$. Given the common use of both effect size types and their mix of advantages and disadvantages we report results for both.

@@

We apply two meta-analytic models, the Hedges and Vevea (1998) random-effects  model and the Hunter and Schmidt (2015) 'bare-bones' random-effects model. We implement these two models as a sensitivity analysis rather than because of an interest in comparing the two in detail. The Hedges and Vevea (HV) model uses inverse variance weight computed as $w_i = 1 / (V_i + T^2)$ (e.g., Borenstein et al., 2010) where $V_i$ is the sampling variance of study $i$ and $T^2$ is the estimated between-studies variance. The Hunter and Schmidt (HS) model weighs studies by sample size $w_i = N_i$. The differences between these weights are relatively minor (Borenstein et al., 2010; Hunter and Schmidt, 2015, p. XX). Although the HV model is typically applied to Fisher z transformed correlations (e.g., Borenstein et al., p. 41-43), we apply it to both transformed and untransformed correlations for the reasons described in the previous section. The HS model typically includes correcting for unreliability or other measurement artifacts. However, we implement it without corrections ('bare-bones'; Hunter & Schmidt, 2015) because this allows us to examine the bias in heterogeneity estimates compared to a perfectly implemented correction of unreliability.

Estimating heterogeneity in both the HV and bare-bones HS model essentially consists of comparing the observed variance in effect sizes with what would be expected from sampling error alone and then ascribing any excess observed variance to heterogeneity (Borenstein et al., 2010, Box 1; Hunter & Schmidt, 2015, p. XX). However, the process for doing so differs somewhat between methods. We apply the Restricted Maximum Likelihood (REML) estimator of heterogeneity when estimating heterogeneity in the Hedges and Vevea (1998) model, because it is a heterogeneity estimator that performs well and is widely implemented (e.g., Viechtbauer, 2005; REF2), and is also the default estimator in the R package metafor. Describing this REML estimation in detail is beyond the scope of this paper and we refer interested readers to the description in Viechtbauer (2005) {this is also what Brannick et al do}. The Hunter and Schmidt (2015) model estimates heterogeneity $\tau^2 = \sigma^2_r - \sigma^2_e$ where  $\sigma^2_r$ is the observed variance across correlations and $\sigma^2_e$ is the average sampling error variance across studies, computed as $ \frac{\Sigma[N_i \sigma^2_{ei}]}{\Sigma N_i}$ (p. 87, 2nd edition) and $\sigma^2_{ei}$ is the usual sampling variance for the Pearson correlation but with $\rho$ replaced with the sample size weighted average correlation $\bar{r} = \frac{\Sigma n_i r_i}{\Sigma n_i}$ instead of $r_i$ as is common. A generalized version of the HS estimator for effect sizes other than correlations can be found in Viechtbauer (2005; 2015) (see ?rma).

>>>>

We focus on correlations as our effect size type in these simulations because correcting them for unreliability is straightforward and because they are the effect size type most commonly corrected in practice. However, unreliability attenuates all types of effect sizes and all can be corrected, either directly (standardized mean differences; Wiernick & Dahlke, 2020) or by first transforming them into correlations. Although effect sizes can generally be transformed into each other (e.g., Borenstein, 2009, p. 46) this should be done with care as it can sometimes affect conclusions (e.g., we found a violation of monotonicity in heterogeneity estimates when converting effect sizes into correlations in a previous project; Olsson-Collentine, 2020, supplement A).

Perhaps the most common way to meta-analyze correlations is to first transform them into Fisher $z$ correlations (Borenstein et al., 2009, p. 41 - 43). The Fisher $z$ transformation is the inverse hyperbolic tangent of the product moment correlation and can be computed as $z = 0.5 \times ln\left(\frac{1+r}{1-r}\right)$ and its variance approximated as $V_z = 1 / (n -3)$. There is controversy in the use of the Fisher $z$ transformation for meta-analysis (e.g., Hunter & Schmidt, 2015, p. XX, ch. 5 "Use of Fisher’s z in Meta-Analysis of Correlations"), largely related to whether Pearson's $r$ or Fisher's $z$ leads to less bias in average correlations (Schulze, 2007). We observe that estimating heterogeneity does not seem to have been the focus of the literature's discussion of the appropriateness of the Fisher $z$ transformation (Brannick et al., 2019; Field, 2005; Hafdahl & Williams, 2009, Hunter & Schmidt, 2015) and that the Fisher $z$ transformation is often applied in practice.

Using Pearson's $r$ for meta-analysis when interested in heterogeneity has the advantage that resulting estimates on this scale are more interpretable that on the Fisher $z$ scale. Although Fisher $z$ correlations can be back-transformed into Pearson's $r$ for interpretation, its heterogeneity cannot. A workaround is to compute a prediction interval and convert this interval onto the product-moment scale (REF). The width of the interval can then be used as an estimate of the heterogeneity. We do not believe this increases interpretability in our case and so prefer to directly estimate the heterogeneity. However, the disadvantage to using Pearson's $r$ when estimating heterogeneity is that 1) the effect size is bounded to {-1, 1} and 2) the sampling variance $$ depends on the effect size. The bounded nature of the Pearson correlation means that large levels of heterogeneity will lead to substantial truncation of effect sizes if the average effect size is also large. That the sampling variance covaries with effect size leads to a small positive bias in heterogeneity estimates (supplement XX). This can be solved by replacing $\rho$ with the average correlation across studies $\bar{r}$ (Hunter & Schmidt, 2015), and removing this covariance is also the primary intention behind the Fisher $z$ transformation. Fisher $z$ correlations also have the advantage that they are not bounded to {-1, 1}. A disadvantage to using Fisher $z$ correlations, in addition to interpretability of heterogeneity estimates, is that they require first backtransforming into product-moment correlations before adding measurement error, and then transforming to Fisher $z$ again for meta-analysis. Especially for large Fisher $z$ values (which can happen if the average effect size and heterogeneity are large) this transformation process may introduce inaccuracies in the estimates. Given the common use of both effect size types and their mix of advantages and disadvantages we report results for both.

@@

We apply two meta-analytic models to ensure that our results are not model-dependent: the Hedges and Vevea (1998) random-effects  model and the Hunter and Schmidt (2015) 'bare-bones' random-effects model. The Hedges and Vevea (HV) model uses inverse variance weight computed as $w_i = 1 / (V_i + T^2)$ (e.g., Borenstein et al., 2010) where $V_i$ is the sampling variance of study $i$ and $T^2$ is the estimated between-studies variance. The Hunter and Schmidt (HS) model weighs studies by sample size $w_i = N_i$. The differences between these weights are relatively minor (Borenstein et al., 2010; Hunter and Schmidt, 2015, p. XX). Although the HV model is typically applied to Fisher z transformed correlations (e.g., Borenstein et al., p. 41-43), we apply it to both transformed and untransformed correlations for the reasons described in the previous section. The HS model typically includes correcting for unreliability or other measurement artifacts. However, we implement it without corrections ('bare-bones'; Hunter & Schmidt, 2015) because this allows us to focus on the bias in heterogeneity estimates rather than the performance of corrections.

Estimating heterogeneity in both the HV and bare-bones HS model essentially consists of comparing the observed variance in effect sizes with what would be expected from sampling error alone and then ascribing any excess observed variance to heterogeneity (Borenstein et al., 2010, Box 1; Hunter & Schmidt, 2015, p. XX). However, the process for doing so differs somewhat between methods. We apply the Restricted Maximum Likelihood (REML) estimator of heterogeneity when estimating heterogeneity in the Hedges and Vevea (1998) model, because it is the generally recommended heterogeneity estimator (Veroniki et al., 2016; Viechtbauer, 2005). Describing the REML estimator in detail is beyond the scope of this paper and we refer interested readers to the description in Viechtbauer (2005). The Hunter and Schmidt (2015) model estimates heterogeneity as $\tau^2 = \sigma^2_r - \sigma^2_e$ where  $\sigma^2_r$ is the observed variance across correlations and $\sigma^2_e$ is the average sampling error variance across studies, computed as $ \frac{\Sigma[N_i \sigma^2_{ei}]}{\Sigma N_i}$ (p. 87, 2nd edition) and $\sigma^2_{ei}$ is the usual sampling variance for the Pearson correlation but with $\rho$ replaced with the sample size weighted average correlation $\bar{r} = \frac{\Sigma n_i r_i}{\Sigma n_i}$ instead of $r_i$ as is common. A generalized version of the HS estimator for effect sizes other than correlations can be found in Viechtbauer (2005; 2015) (see ?rma).
